Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2014-07-17 cordillera-cycle.tex
Polish intro and model set-up.

During the last glacial cycle, glaciers and ice caps of the North American
Cordillera have been more extensive than today. At the Last Glacial
Maximum (LGM), a continuous blanket of ice, the Cordilleran ice sheet\renote{
    In \citep{Seguinot.etal.2014}, The Cryosphere changed all my occurences of
    Cordilleran (Greenland etc) \textbf{I}ce \textbf{S}heet to lowercase
    \textbf{i}ce \textbf{s}heet. Let's keep it so for now for consistency
    within the thesis (kappa + papers).}
\citep{Dawson.1888}, extended from the Alaska Range in the north to the
North Cascades in the south. In addition, it extended offshore, where it calved
into the Pacific Ocean, and merged with the western margin of its much larger
neighbour, the Laurentide ice sheet, east of the Rocky Mountains
(Fig.~\ref{fig:locmap}).

@@

Our understanding of the Cordilleran glaciation history prior to the LGM is
even more fragmentary
    \citep{Barendregt.Irving.1998, Kleman.etal.2010, Rutter.etal.2012},
although it is clear that maximum glaciation of the Cordilleran ice sheet
predates the last glacial cycle
    \citep{Hidy.etal.2013}.
In parts of the Yukon Territory and Alaska, the distribution of tills
    \citep{Turner.etal.2013}
and dated glacial erratics indicate an extensive Marine Oxygen Isotope Stage
(MIS)~4 glaciation
    \citep{Ward.etal.2007, Ward.etal.2008, Briner.Kaufman.2008,
           Stroeven.etal.2010, Stroeven.etal.2014},
yet it is not known whether other regions in the study area were affected.
Landforms in the interior regions include flow sets that are likely
older than the LGM
    \citep[Fig.~2]{Kleman.etal.2010},
but their absolute age remains uncertain.

In contrast, evidence for the deglaciation history of the Cordilleran
ice sheet since the LGM is considerable, albeit mostly at a regional scale.
Geomorphological evidence from south-central British Columbia indicates a rapid
deglaciation, including an early emergence of elevated areas while thin
stagnant ice still covered surrounding lowlands
    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
This model, although credible, may not apply in all areas of the Cordilleran
ice sheet
    \citep{Margold.etal.2013}.
Although solid evidence for late-glacial glacier re-advances have been found in
the Coast, Columbia and Rocky mountains
    \citep{Clague.etal.1997, Friele.Clague.2002, Friele.Clague.2002a,
           Kovanen.2002, Kovanen.Easterbrook.2002, Lakeman.etal.2008,
           Menounos.etal.2008},
it appears to be more sparse than for formerly glaciated regions surrounding
the North Atlantic \aref, thus allowing for considerable uncertainty concerning
the possibility of a regional late glacial cold reversal.

@@

\begin{itemize}
  \item How much ice was locked in the Cordilleran ice sheet during the
    last glacial maximum?
  \item What was the scale of glaciation prior to the last glacial maximum?
  \item Which were its primary dispersal centres? Do they reflect stable or
    ephemeral configurations?
  \item How rapid was the last deglaciation? Did it include late glacial
    stand-stills or re-advances?
\end{itemize}

Although numerical ice sheet modelling has been established as a useful tool to
improve our understanding of the Cordilleran ice sheet
    (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991},
     \citealp{Marshall.etal.2000})
the ubiquitously mountainous
topography of the region has presented two major challenges to its application.
First, only recent development of numerical ice sheet models and underlying
scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2014} has
allowed for high-resolution numerical modelling of glaciers and ice sheets on
mountainous terrain
over millenial time scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the
complex
topography of the North American Cordillera also induces strong gradients in
seasonality and distinct patterns of precipition, thus requiring the use of
high-resolution atmospheric forcing fields as an input to the ice sheet model
\citep{Seguinot.etal.2014}.

Because climate conditions over the last glacial cycle are subject to
considerable uncertainty, our palaeoclimate forcing is a simplistic approach
including temperature and precipitation fields derived from a
present-day atmospheric reanalysis \citep{Mesinger.etal.2006,
Seguinot.etal.2014} supplemented by a lapse-rate correction
and temperature offset time series. The latter are obtained by scaling six
different palaeo-temperature reconstructions from proxy records around the
globe, including two oxygen isotopes records from Greenland ice cores
\citep{Dansgaard.etal.1993, Andersen.etal.2004}, two oxygen isotopes
records from Antarctic ice cores \citep{Petit.etal.1999,Jouzel.etal.2007},
and two alkenone unsaturation index records from Northwest Pacific ocean
sediment cores \citep{Herbert.etal.2001}. We then proceed to compare the model
output to geological evidence and discuss the timing and extent of glaciation
and the patterns of deglaciation.

@@

\todo{Move parameter values into a separate table.}

The simulations presented here were run using the Parallel Ice Sheet Model
(PISM, development version~11b0a7f and stable version~0.6.1), an open source,
finite difference, shallow ice sheet model \citep{PISM-authors.2014}. The model
requires input on basal topography, sea level, geothermal heat flux and
atmospheric\renote{
    Changed here from ``climate'' to ``atmospheric''. Although I have used
    ``climate forcing'' in the previous paper, I now think that ``atmospheric
    forcing'' is more appropriate: ``climate'' may also encompass the oceanic
    component, not really assessed here.}
forcing, and computes the evolution of ice extent
and thickness over time, the thermal and dynamic\renote{
    Thermal = temperature, dynamic = force + movement. If I meant movement
    alone, I would use ``cinematic''. Here we could use ``ice temperatures,
    stress and deformation (i.e. velocities)'', but that becomes a bit clumsy,
    I feel.}
states of the ice sheet, and the associated lithospheric response.

Basal topography is derived from the ETOPO1 combined topography and bathymetry
dataset with a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}. Sea level
is lowered as a function of time based on the Spectral Mapping Project\renote{
    I had no time to check for later products, but if you have suggestions for
    something more up-to-date, I can try to include it in later simulations
    (everything will have to be re-run anyway).}
\citep[SPECMAP;][]{Imbrie.etal.1989} time scale. Geothermal heat flux
is applied as a constant value of 70\,\unit{mW\,m^{-2}} at 3\,km depth
(Sect.~\ref{sec:icedyn}). Atmospheric forcing is provided by a monthly
climatology from the North American Regional Reanalysis
\citep[NARR;][]{Mesinger.etal.2006} perturbated by time-dependent and
lapse-rate temperature corrections (Sect.~\ref{sec:atm}).

@@

Ice sheet dynamics are typically modelled using a combination of internal
deformation and basal sliding. PISM is a~shallow ice sheet model, which implies
that the balance of stresses is approximated based on their predominant
components. The Shallow Shelf Approximation (SSA) is used as a ``sliding law''
for the Shallow Ice Approximation (SIA) by adding both velocity solutions
\citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}\renote{
    Arjen, I don't agree with your comment. Reference to equation number(s)
    show the reader where to look. \citet{Winkelmann.etal.2011} is a rather
    long paper that also treats many other aspects of the model. Their SIA+SSA
    approach builds on \citep{Bueler.Brown.2009}, but it is slightly different.
    It is the one in use in PISM, and in my simulations.}.
Ice rheology depends on temperature and water content through an enthalpy
formulation \citep{Aschwanden.etal.2012}. Surface air temperature derived from
the atmospheric forcing (Sect.~\ref{sec:atm}) provides the upper boundary
condition to the ice enthalpy model. Temperature is computed subglacially to
a~depth of 3\,km\renote{
    Yes, I changed this from 1 to 3\,km as we now model longer period of times
    than in the previous paper. The penetration depth of temperature anomalies
    (i.e. skin depth) depends on the period of temperature oscillations at the
    surface. The longer the period (in our case 100\,kyr), the deeper the
    temperature change. A good way to estimate what this parameter should be is
    to solve the 1-D temperature diffusion equation in the case of periodic
    surface forcing, using rock heat capacity and rock thermal conductivity
    values similar as in the numerical model. This basically forms the basis
    for my choice of the 3\,km value.},
where it is conditioned by a lower boundary geothermal heat flux of
70\,\unit{mW\,m^{-2}}. Although this uniform value does not account for the
high spatial geothermal variability in the region
\citep{Blackwell.Richards.2004}, it is, on average, representative of available
heat flow measurements. In the low-resolution simulations, the vertical grid
consists of to 51~enthalpy layers in the ice sheet and 31~temperature layers in
the bedrock. In the high-resolution simulations, up to 101~ice layers, and
61~bedrock layers are used.

A~pseudo-plastic sliding law \citep{Bueler.Pelt.2014} relates the
bed-parallel shear stress to the sliding velocity. The yield stress, $\tau_c$,
is modelled using the Mohr--Coulomb criterion,
\begin{equation}
   \tau_c = c_0 + N\,\tan{\phi} \,,
\end{equation}
where cohesion, $c_0$, is assumed to be zero. The friction angle, $\phi$,
varies from 15 to 45{\degree}\renote{
    Different from the previous paper, but fairly similar in effect (see
    $\alpha$ and $\delta$ parameter values and corresponding equations in the
    kappa, basal sliding section). These values are is based on a quick
    sensitivity test on Greenland.}.
It is taken to be a~function of modern bed elevation, with lower values at
lower elevations (<200\,m above sea level), thus accounting for a~weakening of
till associated with the presence of marine sediments. Effective pressure, $N$,
is related to the ice overburden stress, $\rho gh$, and the modelled amount of
subglacial water, using a formula derived from laboratory experiments with till
extracted from an Antarctic ice stream \citep{Tulaczyk.etal.2000,
Bueler.Pelt.2014},
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
\end{equation}
where $delta$ is choosed as 0.95, $e_0$ is a measured reference void ratio and
$C_c$ a measured compressibility coefficient. The amount of water at the base,
$W$, varies from zero to $W_{max}=2$\,m, a threshold above which instantaneous
drainage is assumed\renote{
    Regarding a parameter table, it is in the kappa, I am skipping it for now
    but will try to include it soon.}.
Finally, the bedrock topography responds to ice load\renote{
    I disagree with writing ``the isostatic response''. The main point in
    using this model is that computes a non-isostatic response.}
following a bedrock deformation model that includes point-wise isostasy,
elastic lithosphere flexure and viscous mantle deformation in the lower
half-space\renote{
    I had put an emphasis on semi-infinite at this is the key assumption of
    that model, allowing effective solution in Fourier domain. This is the
    whole point of Ed's paper and probably the reason why he picked it for PISM
    But yes, I guess that ``lower half-space'' is fine, too...}
\citep{Lingle.Clark.1985,Bueler.etal.2007}.

% -----------------------------------------------------------------------------
\subsection{Surface mass balance}
% -----------------------------------------------------------------------------

\renote{
    Surface mass balance is the balance of mass fluxes at the surface,
    consisting of ice accumulation and ice ablation, which I introduce straight
    away. I don't know what kind of ``opening sentence'' can be added here.}
Ice surface accumulation and ablation are computed from monthly mean
near-surface air temperature $T_m$, monthly standard deviation of near-surface
air temperature $\sigma$, and monthly precipitation $P_m$, using
a~temperature-index model \citep[e.g.,][]{Hock.2003}. Accumulation is equal to
precipitation when air temperatures are below 0\,\unit{{\degree}C}, and
decreases to zero linearly with temperatures between 0 and
2\,\unit{{\degree}C}. Ablation is computed from the number of positive
degree-days (PDD), defined as the integral of temperatures above
0\,\unit{{\degree}C} in one year.

The PDD computation accounts for stochastic temperature variations by assuming
a normal temperature distribution of standard deviation $\sigma$ aroung the
expected value $T_m$. It is expressed by an error-function formulation
\citep{Calov.Greve.2005},
\begin{equation}
    \label{eqn:calovgreve}
    \mathrm{PDD} = \int_{t_1}^{t_2} \mathrm{d}t
        \left[\frac{\sigma}{\sqrt{2\pi}}
                \exp\left({-\frac{T_{m}^2}{2\sigma^2}}\right)
              + \frac{T_{m}}{2} \, \mathrm{erfc}
                \left(-\frac{T_{m}}{\sqrt{2}\sigma}\right)\right] \,,
\end{equation}
which is numerically approximated using week-long sub-intervals. In order to
account for the effects of spatial and seasonal variations of temperature
variability \citep{Seguinot.2013}, $\sigma$ is computed from daily temperature
values from the North American Regional Reanalysis
\citep[NARR,][]{Mesinger.etal.2006}, after excluding variability associated
with the seasonal cycle itself \citep[cf.][]{Seguinot.Rogozhina.2014}\renote{
    I may add a figure in the kappa if I find time, but I don't think that we
    should do it here. The subject is too marginal for the present study.}.
The ablation model incorporates degree-day factors of
3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia \citep{Shea.etal.2009}.

@@

Atmospheric forcing of the model consists of a present-day monthly climatology
\renote{
    ``Monthly climatology'' is a common term in meteorology to design a set
    of variables aggregated over several decades, and assumed to represent
    some climate state.},
$\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
${\Delta}T_{TS}$, and lapse-rate corrections, ${\Delta}T_{LR}$:
\begin{align}
    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t)
                    + {\Delta}T_{LR}(t, x, y) \,, \\
    P_m(t, x, y) &= P_{m0}(x, y) \,.
\end{align}

The present-day monthly climatology, $\{T_{m0}, P_{m0}\}$, was computed from
near-surface air temperature and precipitation rate fields of the NARR
\citep{Mesinger.etal.2006}, averaged between 1979 and 2000. Modern climate of the
North American Cordillera is characterised by strong geographic variations in
temperature seasonality, timing of the maximum annual precipitation, and
daily temperature variability (Fig.~\ref{fig:atm}).
Our choice of data from the NARR is motivated by the need for an accurate,
high-resolution precipitation forcing, as identified in a previous sensitivity
study \citep{Seguinot.etal.2014}.

Temperature offset time-series, ${\Delta}T_{TS}$, are derived from
palaeo-temperature proxy records from
the Greenland Ice Core Project \citep[GRIP,][]{Dansgaard.etal.1993}, the
North Greenland Ice Core Project \citep[NGRIP,][]{Andersen.etal.2004},
the European Project for Ice Coring in Antarctica \citep[EPICA,][]
{Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and Ocean
Drilling Program (ODP) sites 1012 and 1020, both located off the coast of
California \citep{Herbert.etal.2001}. Palaeo-temperatures from the GRIP and
NGRIP
records were calculated from oxygen isotope (\chem{\delta^{18}O}) measurements
using a quadratic equation \citep{Johnsen.etal.1995},
\begin{equation}
    {\Delta}T_{TS}(t) = -11.88 [\chem{\delta^{18}O}(t)
                               -\chem{\delta^{18}O}(0)]
                        -0.1925[\chem{\delta^{18}O}(t)^2
                                -\chem{\delta^{18}O}(0)^2] \,,
\end{equation}
while temperature reconstructions from Antarctic and oceanic cores were
provided as such. All records were scaled linearly in
order to simulate realistic and comparable ice limits at the last
glacial maximum (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}).

@@

While simulations driven by GRIP and the two Antarctic palaeo-temperature
records attain a last ice volume maximum between 19.5 and 16.9\,ka, those
informed by NGRIP and the two oceanic palaeo-temperature records attain their
maximum ice volume thousands of years earlier. Moreover, the ODP~1012 run
yields a rapid deglaciation of the modelled area prior to 20\,ka. The ODP~1020
simulation predicts an early maximum in ice volume at 29.5\,ka, followed by
slower deglaciation than modelled using the other palaeo-temperature records.
Finally, model runs forced by Antarctic palaeo-temperature records results in a
rapid an uninterrupted deglaciation after the last glacial maximum, whereas the
simulation driven by the GRIP palaeo-temperature record also result in a rapid
deglaciation but in three steps, interrupted by two stand-stills and readvances
of the ice margin.

@@

Despite such differences in the timing of attained volume maxima and mimima,
all the model runs show relatively consistent patterns of glaciation
(Fig.~\ref{fig:snapshots}). As a result of scaling factors applied to the
different palaeo-temperature proxy records (Table~\ref{tab:records}), modelled
ice sheet geometries during the last glacial maximum (MIS~2;
Fig.~\ref{fig:snapshots}, lower panels) appear highly similar from one
simulation to the next. They include a ca. 1500\,km-long central divide
located about 3500\,m\ above sea level (a.s.l.) along the spine of the Rocky
Mountains. The volumes of these modelled last glacial maximum ice sheets are
contained within a close range from 8.24 to 8.66\,m of sea-level equivalent
(s.l.e; Table~\ref{tab:extrema}).
A central ice cap persists in all simulations over the Skeena
Mountains during the MIS~3 ice volume minima (Fig.~\ref{fig:snapshots}, middle
panels). However, the size of the Skeena Mountains ice cap, and the dimensions
of the Cordilleran ice sheet during the MIS~4 glaciation, depend sensitively on
the choice of palaeo-temperature record used to drive the model.

>>>>

During the last glacial cycle, glaciers and ice caps of the North American
Cordillera have been more extensive than today. At the Last Glacial
Maximum (LGM\renote{
    I decided to adopt ``LGM'', which is very common, but not ``CIS'', for the
    sake of readability. This corresponds to abbreviations used in
    \citet{Seguinot.etal.2014} (and now, in the kappa).}),
a continuous blanket of ice, the Cordilleran ice sheet\renote{
    In \citet{Seguinot.etal.2014}, The Cryosphere changed all my occurences of
    Cordilleran (Greenland etc) \textbf{I}ce \textbf{S}heet to lowercase
    \textbf{i}ce \textbf{s}heet. Let's keep it so for now for consistency
    within the thesis (kappa + papers).}
\citep{Dawson.1888}, extended from the Alaska Range in the north to the
North Cascades in the south. In addition, it extended offshore, where it calved
into the Pacific Ocean, and merged with the western margin of its much larger
neighbour, the Laurentide ice sheet, east of the Rocky Mountains
(Fig.~\ref{fig:locmap}).

@@

Our understanding of the Cordilleran glaciation history prior to the LGM is
even more fragmentary
    \citep{Barendregt.Irving.1998, Kleman.etal.2010, Rutter.etal.2012},
although it is clear that maximum glaciation of the Cordilleran ice sheet
predates the last glacial cycle
    \citep{Hidy.etal.2013}.
In parts of the Yukon Territory and Alaska, the distribution of tills
    \citep{Turner.etal.2013}
and dated glacial erratics
    \citep{Ward.etal.2007, Ward.etal.2008, Briner.Kaufman.2008,
           Stroeven.etal.2010, Stroeven.etal.2014}
indicate an extensive Marine Oxygen Isotope Stage (MIS)~4 glaciation,
yet it is not known whether other regions in the study area were affected.
Landforms in the interior regions include flow sets that are likely
older than the LGM
    \citep[Fig.~2]{Kleman.etal.2010},
but their absolute age remains uncertain.

In contrast, evidence for the deglaciation history of the Cordilleran
ice sheet since the LGM is considerable, albeit mostly at a regional scale.
Geomorphological evidence from south-central British Columbia indicates a rapid
deglaciation, including an early emergence of the elevated areas while thin,
stagnant ice still covered the surrounding lowlands
    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
This model, although credible, may not apply in all areas of the Cordilleran
ice sheet
    \citep{Margold.etal.2013}.
Although solid evidence for late-glacial glacier re-advances have been found in
the Coast, Columbia and Rocky mountains
    \citep{Clague.etal.1997, Friele.Clague.2002, Friele.Clague.2002a,
           Kovanen.2002, Kovanen.Easterbrook.2002, Lakeman.etal.2008,
           Menounos.etal.2008},
it appears to be more sparse than for formerly glaciated regions surrounding
the North Atlantic \aref\footnote{
    Do we need to add a couple of refs. from Britain, Scandinavia, the Alps,
    Eastern America here? Or a review paper? Could yuo help me with this?},
thus allowing for considerable uncertainty concerning
the possibility of a regional late glacial cold reversal.

@@

\begin{itemize}
  \item How much ice was locked in the Cordilleran ice sheet during the
    last glacial maximum?
  \item What was the scale of glaciation prior to the last glacial maximum?
  \item Which were the primary dispersal centres? Do they reflect stable or
    ephemeral configurations?
  \item How rapid was the last deglaciation? Did it include late glacial
    standstills or readvances?
\end{itemize}

Although numerical ice sheet modelling has been established as a useful tool to
improve our understanding of the Cordilleran ice sheet
    (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991},
     \citealp{Marshall.etal.2000}),
the ubiquitously mountainous
topography of the region has presented two major challenges to its application.
First, only recent development of numerical ice sheet models and underlying
scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2014} has
allowed for high-resolution numerical modelling of glaciers and ice sheets on
mountainous terrain
over millenial time scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the
complex
topography of the North American Cordillera also induces strong gradients in
seasonality and distinct patterns of precipition, thus requiring the use of
high-resolution atmospheric forcing fields as an input to the ice sheet model
\citep{Seguinot.etal.2014}.

Because climate conditions over the last glacial cycle are subject to
considerable uncertainty, our palaeo-climate forcing is a simplistic approach
including temperature and precipitation fields derived from a
present-day atmospheric reanalysis \citep{Mesinger.etal.2006,
Seguinot.etal.2014} supplemented by lapse-rate corrections
and temperature offset time series. The latter are obtained by scaling six
different palaeo-temperature reconstructions from proxy records around the
globe, including two oxygen isotope records from Greenland ice cores
\citep{Dansgaard.etal.1993, Andersen.etal.2004}, two oxygen isotope
records from Antarctic ice cores \citep{Petit.etal.1999,Jouzel.etal.2007},
and two alkenone unsaturation index records from Northwest Pacific ocean
sediment cores \citep{Herbert.etal.2001}. We then proceed to compare the model
output to geological evidence and discuss the timing and extent of glaciation
and the patterns of deglaciation.

@@

The simulations presented here were run using the Parallel Ice Sheet Model
(PISM, development version~11b0a7f and stable version~0.6.1), an open source,
finite difference, shallow ice sheet model \citep{PISM-authors.2014}. The model
requires input on basal topography, sea level, geothermal heat flux and
atmospheric\renote{
    Changed here from ``climate'' to ``atmospheric''. Although I have used
    ``climate forcing'' in the previous paper, I now think that ``atmospheric
    forcing'' is more appropriate: ``climate'' may also encompass the oceanic
    component, which is not assessed here.}
forcing. It computes the evolution of ice extent
and thickness over time, the thermal and dynamic\renote{
    Thermal = temperature, dynamic = force + movement. If I meant movement
    alone, I would use ``cinematic''. Here we could use ``ice temperatures,
    stress and deformation (i.e. velocities)'', but that becomes a bit clumsy,
    I feel.}
states of the ice sheet, and the associated lithospheric response.

Basal topography is derived from the ETOPO1 combined topography and bathymetry
dataset with a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}. Sea level
is lowered as a function of time based on the Spectral Mapping Project\renote{
    I had no time to check for later products, but if you have suggestions for
    something more up-to-date, I can try to include it in later simulations
    (everything will have to be re-run anyway).}
\citep[SPECMAP;][]{Imbrie.etal.1989} time scale. Geothermal heat flux
is applied as a constant value of 70\,\unit{mW\,m^{-2}} at 3\,km depth
(Sect.~\ref{sec:icedyn}). Surface mass balance is computed using a positive
degree-day (PDD) model (Sect.~\ref{sec:surface}). Atmospheric forcing is
provided by a monthly climatology from the North American Regional Reanalysis
\citep[NARR;][]{Mesinger.etal.2006} perturbated by time-dependent and
lapse-rate temperature corrections (Sect.~\ref{sec:atm}).

@@

Ice sheet dynamics are typically modelled using a combination of internal
deformation and basal sliding. PISM is a~shallow ice sheet model, which implies
that the balance of stresses is approximated based on their predominant
components. The Shallow Shelf Approximation (SSA) is used as a ``sliding law''
for the Shallow Ice Approximation (SIA) by adding velocity solutions of the
two approximation
\citep[Eqns.~7--9 and 15]{Bueler.Brown.2009, Winkelmann.etal.2011}\renote{
    Arjen, I don't agree with your comment. Reference to equation number(s)
    show the reader where to look. \citet{Winkelmann.etal.2011} is a rather
    long paper that also treats many other aspects of the model. Their SIA+SSA
    approach builds on \citep{Bueler.Brown.2009}, but it is slightly different.
    It is the one in use in PISM, and in my simulations.}.
Ice rheology depends on temperature and water content through an enthalpy
formulation \citep{Aschwanden.etal.2012}. Surface air temperature derived from
the atmospheric forcing (Sect.~\ref{sec:atm}) provides the upper boundary
condition to the ice enthalpy model. Temperature is computed subglacially to
a~depth of 3\,km\renote{
    Yes, I changed this from 1 to 3\,km as we now model longer period of times
    than in the previous paper. The penetration depth of temperature anomalies
    (i.e. skin depth) depends on the period of temperature oscillations at the
    surface. The longer the period (in our case 100\,kyr), the deeper the
    temperature change. A good way to estimate what this parameter should be is
    to solve the 1-D temperature diffusion equation in the case of periodic
    surface forcing, using rock heat capacity and rock thermal conductivity
    values similar as in the numerical model. This basically forms the basis
    for my choice of the 3\,km value.},
where it is conditioned by a lower boundary geothermal heat flux of
70\,\unit{mW\,m^{-2}}. Although this uniform value does not account for the
high spatial geothermal variability in the region
\citep{Blackwell.Richards.2004}, it is, on average, representative of available
heat flow measurements. In the low-resolution simulations, the vertical grid
consists of up to 51~enthalpy layers in the ice sheet and 31~temperature layers
in the bedrock. In the high-resolution simulations, up to 101~ice layers, and
61~bedrock layers are used.

A~pseudo-plastic sliding law \citep{Bueler.Pelt.2014} relates the
bed-parallel shear stresses to the sliding velocity\footnote{
    Should I pull in this equation, too?}.
The yield stress, $\tau_c$,
is modelled using the Mohr--Coulomb criterion,
\begin{equation}
   \tau_c = c_0 + N\,\tan{\phi} \,,
\end{equation}
where cohesion, $c_0$, is assumed to be zero. The friction angle, $\phi$,
varies from 15 to 45{\degree}\renote{
    Different from the previous paper, but fairly similar in effect (see
    $\alpha$ and $\delta$ parameter values and corresponding equations in the
    kappa, basal sliding section). These values are is based on a quick
    sensitivity test on Greenland.}.
It is taken to be a piecewise-linear function of modern bed elevation, with
the lowest value below modern sea level (0\,m above sea level, m~a.s.l) and the
highest value above the approximate elevation of highest shorelines
(200\,m~a.s.l), thus accounting for a~weakening of
till associated with the presence of marine sediments. Effective pressure, $N$,
is related to the ice overburden stress, $\rho gh$, and the modelled amount of
subglacial water, using a formula derived from laboratory experiments with till
extracted from an Antarctic ice stream \citep{Tulaczyk.etal.2000,
Bueler.Pelt.2014},
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
\end{equation}
where $delta$ is chosen as 0.02, $e_0$ is a measured reference void ratio and
$C_c$ a measured compressibility coefficient (Table~\ref{tab:params}). The
amount of water at the base, $W$, varies from zero to $W_{max}=2$\,m, a
threshold above which instantaneous drainage is assumed.
Finally, the bedrock topography responds to ice load\renote{
    I disagree with writing ``the isostatic response''. The main point in
    using this model is that computes a non-isostatic response.}
following a bedrock deformation model that includes point-wise isostasy,
elastic lithosphere flexure and viscous mantle deformation in the lower
half-space\renote{
    I had put an emphasis on semi-infinite as this is, from a mathematical
    point of view, the key assumption of that model, allowing effective
    solution in Fourier domain. This is the whole point of Ed's paper and
    probably the reason why he picked it for PISM. But yes, I guess that
    ``lower half-space'' is fine, too...}
\citep[Table~\ref{tab:params};][]{Lingle.Clark.1985,Bueler.etal.2007}.

% -----------------------------------------------------------------------------
\subsection{Surface mass balance}
\label{sec:surface}
% -----------------------------------------------------------------------------

\renote{
    Surface mass balance is the balance of mass fluxes at the surface,
    consisting of ice accumulation and ice ablation, which I introduce straight
    away. I don't see what kind of ``opening sentence'' can be added here.}
Ice surface accumulation and ablation are computed from monthly mean
near-surface air temperature, $T_m$, monthly standard deviation of near-surface
air temperature, $\sigma$, and monthly precipitation, $P_m$, using
a~temperature-index model \citep[e.g.,][]{Hock.2003}. Accumulation is equal to
precipitation when air temperatures are below 0\,\unit{{\degree}C}, and
decreases to zero linearly with temperatures between 0 and
2\,\unit{{\degree}C}. Ablation is computed from the number of positive
degree-days (PDD), defined as the integral of temperatures above
0\,\unit{{\degree}C} in one year.

The PDD computation accounts for stochastic temperature variations by assuming
a normal temperature distribution of standard deviation $\sigma$ aroung the
expected value $T_m$. It is expressed by an error-function formulation
\citep{Calov.Greve.2005},
\begin{equation}
    \label{eqn:calovgreve}
    \mathrm{PDD} = \int_{t_1}^{t_2} \mathrm{d}t
        \left[\frac{\sigma}{\sqrt{2\pi}}
                \exp\left({-\frac{T_{m}^2}{2\sigma^2}}\right)
              + \frac{T_{m}}{2} \, \mathrm{erfc}
                \left(-\frac{T_{m}}{\sqrt{2}\sigma}\right)\right] \,,
\end{equation}
which is numerically approximated using week-long sub-intervals. In order to
account for the effects of spatial and seasonal variations of temperature
variability \citep{Seguinot.2013}, $\sigma$ is computed from daily temperature
values from the North American Regional Reanalysis
\citep[NARR,][]{Mesinger.etal.2006}, after excluding variability associated
with the seasonal cycle itself \citep[cf.][]{Seguinot.Rogozhina.2014}\renote{
    I may add a figure in the kappa if I find time, but I don't think that we
    should do it here. The subject is too marginal for the present study.}.
Degree-day factors for snow and ice melt are derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia
\citep[Table~\ref{tab:params};][]{Shea.etal.2009}.

@@

Atmospheric forcing of the model consists of a present-day monthly climatology%
\renote{
    ``Climatology'' is a common term in meteorology, used to design a set
    of variables aggregated over several decades, and assumed to represent
    some climate state.},
$\{T_{m0}, P_{m0}\}$, where temperatures are modified by offset time series,
${\Delta}T_{TS}$, and lapse-rate corrections, ${\Delta}T_{LR}$:
\begin{align}
    T_m(t, x, y) &= T_{m0}(x, y) + {\Delta}T_{TS}(t)
                    + {\Delta}T_{LR}(t, x, y) \,, \\
    P_m(t, x, y) &= P_{m0}(x, y) \,.
\end{align}

The present-day monthly climatology, $\{T_{m0}, P_{m0}\}$, was computed from
near-surface air temperature and precipitation rate fields from the NARR
\citep{Mesinger.etal.2006}, averaged between 1979 and 2000. Modern climate of the
North American Cordillera is characterised by strong geographic variations in
temperature seasonality, timing of the maximum annual precipitation, and
daily temperature variability (Fig.~\ref{fig:atm}).
Our choice of data from the NARR is motivated by the need for an accurate,
high-resolution precipitation forcing, as identified in a previous sensitivity
study \citep{Seguinot.etal.2014}.

Temperature offset time-series, ${\Delta}T_{TS}$, are derived from
palaeo-temperature proxy records from
the Greenland Ice Core Project \citep[GRIP,][]{Dansgaard.etal.1993}, the
North Greenland Ice Core Project \citep[NGRIP,][]{Andersen.etal.2004},
the European Project for Ice Coring in Antarctica \citep[EPICA,][]
{Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and Ocean
Drilling Program (ODP) sites 1012 and 1020, both located off the coast of
California \citep{Herbert.etal.2001}. Palaeo-temperatures from the GRIP and
NGRIP
records were calculated from oxygen isotope (\chem{\delta^{18}O}) measurements
using a quadratic equation \citep{Johnsen.etal.1995},
\begin{equation}
    {\Delta}T_{TS}(t) = -11.88 [\chem{\delta^{18}O}(t)
                               -\chem{\delta^{18}O}(0)]
                        -0.1925[\chem{\delta^{18}O}(t)^2
                                -\chem{\delta^{18}O}(0)^2] \,,
\end{equation}
while temperature reconstructions from Antarctic and oceanic cores were
provided as such. All records were scaled linearly (Table~\ref{tab:records}) in
order to simulate realistic and comparable ice extents at the LGM
(Fig.~\ref{fig:timeseries}).

@@

While simulations driven by GRIP and the two Antarctic palaeo-temperature
records attain a last ice volume maximum between 19.5 and 16.9\,ka, those
informed by NGRIP and the two oceanic palaeo-temperature records attain their
maximum ice volume thousands of years earlier. Moreover, the ODP~1012 run
yields a rapid deglaciation of the modelled area prior to 20\,ka. The ODP~1020
simulation predicts an early maximum in ice volume at 29.5\,ka, followed by
slower deglaciation than modelled using the other palaeo-temperature records.
Finally, model runs forced by Antarctic palaeo-temperature records results in a
rapid an uninterrupted deglaciation after the last glacial maximum, whereas the
simulation driven by the GRIP palaeo-temperature record also result in a rapid
deglaciation but in three steps, interrupted by two standstills and readvances
of the ice margin.

@@

Despite such differences in the timing of attained volume maxima and mimima,
all the model runs show relatively consistent patterns of glaciation
(Fig.~\ref{fig:snapshots}). As a result of scaling factors applied to the
different palaeo-temperature proxy records (Table~\ref{tab:records}), modelled
ice sheet geometries during the last glacial maximum (MIS~2;
Fig.~\ref{fig:snapshots}, lower panels) appear highly similar from one
simulation to the next. They include a ca. 1500\,km-long central divide
located about 3500\,m~a.s.l. along the spine of the Rocky
Mountains. The volumes of these modelled last glacial maximum ice sheets are
contained within a close range from 8.24 to 8.66\,m of sea-level equivalent
(s.l.e; Table~\ref{tab:extrema}).
A central ice cap persists in all simulations over the Skeena
Mountains during the MIS~3 ice volume minima (Fig.~\ref{fig:snapshots}, middle
panels). However, the size of the Skeena Mountains ice cap, and the dimensions
of the Cordilleran ice sheet during the MIS~4 glaciation, depend sensitively on
the choice of palaeo-temperature record used to drive the model.
